\vspace{3cm} \Huge{\futhark}\\ \Large{Enhancing performance by rewriting kernels from Python/OpenCL to Futhark}

### 2017-11-09 report/draft.tex
asd

\clearpage\maketitle
\thispagestyle{empty}

\section*{Abstract}
This report describes the design and implementation of a Futhark extension for
the modelling engine \sasmodels.
\sasmodels{} performs small angle scattering calculations for the \sasview{}
modelling suite, by sending it's input to a range of calculation models
implemented in Python and OpenCL.\\\\

%%% whaaaat

@@

\section{Introduction}
\subsection{Goal and motivation}
\sasview{} is a Small Angle Scattering (SAS) analysis package for the
analysis of 1D and 2D scattering data directly in inverse space.
As \sasview{} is given data, it performs its analyses by running the
input data through theory models provided by the standalone module
\sasmodels{}.
\\\\
The \sasmodels{} models utilizes a set of kernels, which are implemented
in both Python and OpenCL. However, benchmark
comparisons\footnote{https://futhark-lang.org/performance.html}
between the parallel algorithm library \textit{Thrust}
(implemented in C++/CUDA), and the array programming language Futhark, shows
that there might severe speed improvements to gain by running these model
calculations in Futhark, instead of in OpenCL as they are now.
\\\\
The goal of this project is to explore the potential speed gains obtained by
rewriting, and testing the resulting speed improvements of, a subset of the
\sasmodels{} models.
The subset will be chosen so that there are at least
one ``simple'' model, one model of intermediate complexity, and one complex
model in it.
\\\\
If rewriting the models results in a significant\footnote{I.e. a speed increase
  in order of at least a magnitude}, speedup, it will reasonable to believe that 
  the rest of the models in \sasmodels{} can be rewritten for a similar
  speedup.

\section{Demonstration}
\begin{mdframed}[frametitle={Terminology:}]
  The work of this project has been spread out across multiple parts
  of the \sasmodels{} program. Here is a list of the distinct parts of \sasmodels
  that has been worked with:
  \begin{itemize}
    \item \texttt{kernelfut}: The \sasmodels{} Futhark kernel module that wraps 
    \sasmodels{} input data and sends it to be calculated in the actual Futhark 
    kernel. \texttt{kernelfut} is an alternative to \texttt{kernelpy} and
    \texttt{kernelcl} which are already available in \sasmodels.
    The final kernel object is FutKernel.

@@

\subsection{Describing \sasmodels' models}
\begin{itemize}
  \item a function definition \texttt{I(q)}, which defines scattered intensity 
  of the scattering vector \texttt{q}.

@@

Additionally, the model info file also always contains the short-name, 
the full title, a category, and a mathematical \footnote{a representation 
that compiles to a meaningful mathematical expression if used in a LaTeX math 
environment} string representation of the scattering function.
\\\\
As the scattering function \texttt{I(q)} and its parameters are the only 
two things that are strictly necessary in a model definition, the developer
has the option to leave the model info defined with just these six properties.
\\\\
\sasmodels{} has built-in default definitions for $I(q_x,q_y)$ if working
with two-dimensional data, and other functions such as \texttt{form\_volume(q)}
for working with voluminous scatterings.
\\\\
Therefore, it is optional for the developer to define a custom definition
of $I(q_x, q_y)$ or other default functions.
An example of this can be seen in figure \ref{fig:linemodel}.
A more complex model can be seen in appendix \ref{fig:core_shell}

@@

\begin{mdframed}[
  frametitle={Why does Futhark perform faster than Python?},
  nobreak=true
  ]
These comparisons are between Python/Numpy- and Futhark models.
Although Numpy can execute both of the models as vectorized calculations, 
it is still vastly outperformed by Futhark, which can run hundreds of threads 
simultaneously on the GPU. In comparison, Numpy runs on the CPU, which rarely 
runs more than 10-20 threads simultaneously.
\\
For models like broad\_peak, Futhark gains a significant speedup by 
hoisting\cite[sec 2.3]{fhpc17} the allocation of variables 
(fig. \ref{fig:broadpeakmodel_futhark}) outside of
the function call, so they are allocated as constants in the \foriq{} 
\texttt{run\_kernel} function, right before mapping Iq.Iq over the q input.
\\\\
Hoisting is not uncommon for compilers to do at all, but it does make a 
difference for \sasmodels{}.
The compiler moves the number of variable allocations from
$O(\text{length of q vector } \cdot \text{number of parameters})$ to 
$O(\text{number of parameters})$.
\end{mdframed}

@@

While it was expected to see Futhark outperform the strictly CPU based 
computations performed by Numpy in the Python models, it was more uncertain
how much (or even whether) Futhark would outperform OpenCL.
\\\\
However, data shows that Futhark does perform better. The increased performance 
in figure \ref{fig:dab-graph} is not really significant until we reach very high
n, which can be attributed to the simplicity of the model.
\\
With such simple models, there are not many things that Futhark can do smarter 
than the normal OpenCL compiler, besides the gains that are possible from
the aforementioned hoisting.
\\\\
However, when we work with the rewritten core shell parallelepiped, we see huge
speedups even at n=100. This speedup comes from the Futhark compiler's 
nested \texttt{map} optimizations: the Iq function is mapped
over the array of q values, and the Iq function body is mainly a nested 
map-map-reduce-reduce, which makes the final nested function a 
map-map-map-reduce-reduce. 
\\
As this construct can be flattened 
and parallelised\cite[sec. 5]{pldi17}, the Futhark compiler ensures that the
calculations within the maps over the Gauss arrays and q values are coalesced,
we are guaranteed efficient memory access.
For the original OpenCL implementation, we risk that different kernels are
accessing memory locations that are physically far apart, so that the GPU
has to handle the memory accesses individually, instead of being able to 
combine them.
\end{mdframed}

\subsection{Correctness of Futhark results}
\label{sec:correctness}
There has been one small issue with the correctness of the Futhark models.
As the Futhark models are generic by design, and instantiated with the 
programmer's floating point type of choice, we can run the Futhark kernels
with any precision we want.
\\\\
However, some of the original \sasmodels{} models have Iq functions that are 
hardcoded to use double precision floats in places, and even sometimes mixes
doubles and singles. 
As these models plays fast and loose with the precision type, the Futhark models
will inevitably end up showing some total difference between it's results and 
the results from the built in functions.
\\\\
However, several \sasview{} users have assured me, that it is not really a 
problem. The worst examples of these differences has been accumulated errors of
between 10e-6 and 10e-7, which is well within the error range accepted by 
the \sasview{} users.
\section{Discussion, future work}
\label{sec:discussion}
The performance tests in \ref{performance} unequivocally shows, that
there are speedups to gain by running Futhark kernels in favor of Python or
OpenCL kernels.
In general, Futhark starts surpassing the performance of competing kernels, when
sample sizes rise above n=INDSÆT N. However, these gains are currently
irrelevant for the most \sasview{} use cases, as almost all
experiments are carried out on sample sizes of between 100-500.
However, meetings with external project owner Wojciech Potrzebowski lead to the
suggestion that simple bundling of \sasview{} experiments would enable us
to leverage the Futhark speed improvements directly.
\\\\
In example, one could imagine running an experiment that continuously sends
sensor data for processing in \sasview{} / \sasmodels.
Instead of running an experiment for each sample\footnote{of size n=100-500}
from the stream, one could let \sasmodels{} bundle these input samples in packages
of 10-50 samples
instead, and calculate these collectively to leverage the speedups that are
apparent on Futhark calculations on samples of larger sizes.
\\\\
Please note, that sample sizes might have an upper bound, depending on the
memory available on the workstations running the experiments.

@@

\subsection{Usability of Futhark models}
It is relevant to discuss how useful Futhark models are to the \sasview{} user
groups. For the users who merely wants to use the models, it is not necessary
to understand the implementation itself at all, as the user facing part of
the model is just the model info file, which is just the same as the model infos
for Python and OpenCL models \ref{sec:using-futhark}.
\\\\
However, for the model developer, it is now also necessary to have at least a
rudimentary understanding of functional programming.
In example, developers will have to learn how to use maps in
places where they would usually use loops.
\\\\
Furthermore, the current solution requires severe amounts of boilerplate code to
be included in the futhark model file.
This boilerplate is caused by the usage of Futhark's module system, and
therefore not strictly part of defining i.e. the \iq function of a model.
The boilerplate might become a distraction for the model developer, so
there might be a smaller educational challenge to solve here as well.
\\\\
\section{Conclusion}
To answer the questions raised in this project definition, it was first
necessary to implement a \sasmodels{} kernel for Futhark models. This was done
successfully and in a very non-destructive way in the already existing
\sasmodels{} code base.
\\\\
The kernel itself (\texttt{kernelfut.py}) depends on the existing code base, but
the total change of the original code is two lines: an extra case in the
switch statement that decides which kernel type to run for a given model.
\\\\
For Futhark model performances increases themselves, the results in
sec \ref{performance}, tells us that there are performance increase of between
200- and 2000 percent, depending on the original model's language and
complexity.
\\\\
These gains are only realised when the sample size for a given run is high
enough, but section \ref{sec:discussion} has proposals for \sasmodels{} extensions
that could help \sasmodels{} users to reach these sample sizes, by bundling
experiments.
\\\\
\begin{mdframed}[
  frametitle={Presentation at \sasview{} Code Camp},
  nobreak=true
  ]
Together with my supervisor and Futhark designer Troels Henriksen,
I was invited for an hour-long presentation and discussion of \futhark{} at the
yearly \textit{SasView Code Camp}, held at the Copenhagen Bio Science Park.\\\\
There was about 20 \sasview{} users and developers attending at the presentation,
which was about the definition and implementation of \futhark{}, along with
a presentation of the result performance boosts for selected models.\\\\
The project was well-received by the audience, and there was an eager discussion
of whether and how it could be possible to really exploit the Futhark
optimizations.\\\\
The primary suggestion was to leverage Futhark to find best-fitting models,
as described in \ref{sec:best-match}, but bundling continuous input from
a sensor to exploit Futhark speed ups at high n was also a feature that many
of the attendants expressed an interest in.
\end{mdframed}

@@

\begin{figure}
  \lstinputlisting[linerange={13-53}]{../sasmodels/sasmodels/models/futharks/header/for_iq.fut}
  \label{fig:kernelfut_loop}
  \caption{The Futhark kernel calculation loop, aping fig \ref{fig:kernelpy_loop}}
\end{figure}

\begin{figure}
  \lstinputlisting[linerange={541-588}, language=Python]{../sasmodels/sasmodels/kernelcl.py}
  \label{fig:kernelcl_call}
  \caption{Calling OpenCL kernel}
\end{figure}

@@

\end{document}

>>>>

\clearpage\maketitle
\thispagestyle{empty}
\clearpage
\section*{Abstract}
\sasmodels{} performs small angle scattering calculations for the \sasview{}
modelling suite, by sending it's input to a range of calculation models
implemented in Python and OpenCL. 
\\
A Futhark module was written and added to the 
existing \sasmodels{} codebase, to test whether Futhark could outperform the 
already existing model languages used in \sasmodels{}. The Futhark models
were performance tested against a range of the original models.
All following results showed a speed increase of between 200 and 2000\%, 
all without loss of precision.  Speedups were however only apparent at 
unrealistically large input sizes.
\\
Future \sasmodels{} extensions are then suggested, so these speed increases can be 
leveraged in daily \sasview{} usage.

@@

\section{Introduction}
\sasview{} is a Small Angle Scattering (SAS) analysis package for the
analysis of 1D and 2D scattering data directly in inverse space.
When \sasview{} is given input data, it performs its analyses by running the
input data through theory models provided by the standalone module
\sasmodels{}.
\\\\
The \sasmodels{} models utilizes a set of kernels, which are implemented
in both Python and OpenCL. However, benchmark
comparisons\cite{futhark}
between the parallel algorithm library \textit{Thrust}
(implemented in C++/CUDA), and the array programming language Futhark, shows
that there might severe speed improvements to gain by running these model
calculations in Futhark, instead of in OpenCL as they are now.
\\\\
The goal of this project is to explore the potential speed gains obtained by
rewriting, and testing the resulting speed improvements of, a subset of the
\sasmodels{} models.
The subset will be chosen so that there are for each of Python and OpenCL 
one ``simple'' model, and a complex one.
\\\\
If rewriting the models results in a significant\footnote{I.e. a speed increase
  in order of at least 100\%}, speedup, it will reasonable to believe that 
  the rest of the models in \sasmodels{} can be rewritten for a similar
  speedup.

\section{Demonstration}
\begin{mdframed}[
  frametitle={Terminology:},
  nobreak=true
  ]
  The work of this project has been spread out across multiple parts
  of the \sasmodels{} program. Here is a list of the distinct parts of 
  \sasmodels{} that has been worked with:
  \begin{itemize}
    \item \texttt{kernelfut}: The \sasmodels{} Futhark kernel module that wraps 
    \sasmodels{} input data and sends it to be calculated in the actual Futhark 
    kernel. \texttt{kernelfut} is an alternative to \texttt{kernelpy} and
    \texttt{kernelcl} which are already available in \sasmodels.

@@

All \sasmodels{} models are described in model info file. These model info 
files has at least two properties:
\begin{itemize}
  \item a function definition \texttt{I(q)}, which defines scattered intensity 
  of the scattering vector \texttt{q}.

@@

Additionally, the model info file also always contains the short-name, 
the full title, a category, and a mathematical \footnote{a representation 
that compiles to a meaningful mathematical expression if used in a LaTeX math 
environment} string representation of the scattering function.
\\\\
As the scattering function \texttt{I(q)} and its parameters are the only 
two things that are strictly necessary in a model definition, the developer
has the option to leave the model info defined with just these six properties.
\\\\
\sasmodels{} has built-in default definitions for $I(q_x,q_y)$ if working
with two-dimensional data, and other functions such as \texttt{form\_volume(q)}
for working with voluminous scatterings.
\\\\
Therefore, it is optional for the developer to define a custom definition
of $I(q_x, q_y)$ or other default functions.
An example of this can be seen in fig. \ref{fig:linemodel}.
A more complex model can be seen in fig. \ref{fig:core_shell}

@@

\begin{mdframed}[
  frametitle={Why does Futhark perform faster than Python?},
  nobreak=true
  ]
These comparisons are between Python/Numpy- and Futhark models.
Although Numpy can execute both of the models as vectorized calculations, 
it is still vastly outperformed by Futhark, which can run hundreds of threads 
simultaneously on the GPU. In comparison, Numpy runs on the CPU, which rarely 
runs more than 10-20 threads simultaneously.
\\
For models like broad\_peak, Futhark gains a significant speedup by 
hoisting\cite[sec 2.3]{pldi17} the allocation of variables 
(fig. \ref{fig:broadpeakmodel_futhark}) outside of
the function call, so they are allocated as constants in the \foriq{} 
\texttt{run\_kernel} function, right before mapping Iq.Iq over the q input.
\\\\
Hoisting is not uncommon for compilers to do at all, but it does make a 
difference for \sasmodels{}.
The compiler moves the number of variable allocations from
$O(\text{length of q vector } \cdot \text{number of parameters})$ to 
$O(\text{number of parameters})$.
\end{mdframed}

@@

While it was expected to see Futhark outperform the strictly CPU based 
computations performed by Numpy in the Python models, it was more uncertain
how much (or even whether) Futhark would outperform OpenCL.
\\\\
However, data shows that Futhark does perform better. The increased performance 
in figure \ref{fig:dab-graph} is not really significant until we reach very high
n, which can be attributed to the simplicity of the model.
\\
With such simple models, there are not many things that Futhark can do smarter 
than the normal OpenCL compiler, besides the gains that are possible from
the aforementioned hoisting.
\\\\
However, when we work with the rewritten core shell parallelepiped, we see huge
speedups even at n=100. This speedup comes from the Futhark compiler's 
nested \texttt{map} optimizations: the Iq function is mapped
over the array of q values, and the Iq function body is mainly a nested 
map-map-reduce-reduce, which makes the final nested function a 
map-map-map-reduce-reduce. 
\\
As this construct can be flattened and parallelised\cite[sec. 5]{pldi17}, 
we are can split our workload into many more tasks than the original OpenCL 
implementation, which only has nq tasks. However, every single of these tasks
consists of complex for-loop nests, and plenty of variable assignments.
\\\\
Furthermore, the Futhark compiler takes care to ensure that the calculations 
over the Gauss arrays and q values are coalesced, so we are given 
efficient memory access.
\\
For the original OpenCL implementation, we risk that different kernels are
accessing memory locations that are physically far apart, so that the GPU
has to handle the memory accesses individually, instead of being able to 
combine them.
\end{mdframed}

\subsection{Correctness of Futhark results}
\label{sec:correctness}
There has been one small issue with the correctness of the Futhark models.
As the Futhark models are generic by design, and instantiated with the 
programmer's floating point type of choice, we can run the Futhark kernels
with any precision we want.
\\\\
However, some of the original \sasmodels{} models have Iq functions that are 
hardcoded to use double precision floats in places, and even sometimes mixes
doubles and singles. 
As these models plays fast and loose with the precision type, the Futhark models
will inevitably end up showing some total difference between it's results and 
the results from the built in functions.
\\\\
However, several \sasview{} users have assured me, that it is not really a 
problem. The worst examples of these differences has been accumulated errors of
between 10e-6 and 10e-7, which is well within the error range accepted by 
the \sasview{} users.
\section{Discussion, future work}
\label{sec:discussion}
The performance tests in \ref{sec:performance} unequivocally shows, that
there are speedups to gain by running Futhark kernels in favor of Python or
OpenCL kernels.
In general, Futhark starts surpassing the performance of competing kernels, when
sample sizes rise above n=INDSÆT N. However, these gains are currently
irrelevant for the most \sasview{} use cases, as almost all
experiments are carried out on sample sizes of between 100-500.
However, meetings with external project owner Wojciech Potrzebowski lead to the
suggestion that simple bundling of \sasview{} experiments would enable us
to leverage the Futhark speed improvements directly.
\\\\
In example, one could imagine running an experiment that continuously sends
sensor data for processing in \sasview{} / \sasmodels.
Instead of running an experiment for each sample\footnote{of size n=100-500}
from the stream, one could let \sasmodels{} bundle these input samples in packages
of 10-50 samples
instead, and calculate these collectively to leverage the speedups that are
apparent on Futhark calculations on samples of larger sizes.
\\\\
Please note, that sample sizes might have an upper bound, depending on the
memory available on the workstations running the experiments.

@@

\subsection{Usability of Futhark models}
It is relevant to discuss how useful Futhark models are to the \sasview{} user
groups. For the users who merely wants to use the models, it is not necessary
to understand the implementation itself at all, as the user facing part of
the model is just the model info file, which is just the same as the model infos
for Python and OpenCL models \ref{sec:using-futhark}.
\\\\
However, for the model developer, it is now also necessary to have at least a
rudimentary understanding of functional programming.
In example, developers will have to learn how to use maps in
places where they would usually use loops.
\\\\
Furthermore, the current solution requires severe amounts of boilerplate code to
be included in the futhark model file.
This boilerplate is caused by the usage of Futhark's module system, and
therefore not strictly part of defining i.e. the \iq function of a model.
The boilerplate might become a distraction for the model developer, so
there might be a smaller educational challenge to solve here as well.
\\\\
\section{Conclusion}
To answer the questions raised in this project definition, it was first
necessary to implement a \sasmodels{} kernel for Futhark models. This was done
successfully and in a very non-destructive way in the already existing
\sasmodels{} code base.
\\\\
The kernel itself (\texttt{kernelfut.py}) depends on the existing code base, but
the total change of the original code is two lines: an extra case in the
switch statement that decides which kernel type to run for a given model.
\\\\
For Futhark model performances increases themselves, the results in
sec \ref{sec:performance}, tells us that there are performance increase of between
200- and 2000 percent, depending on the original model's language and
complexity.
\\\\
These gains are only realised when the sample size for a given run is high
enough, but section \ref{sec:discussion} has proposals for \sasmodels{} extensions
that could help \sasmodels{} users to reach these sample sizes, by bundling
experiments.
\\\\
\begin{mdframed}[
  frametitle={Presentation at \sasview{} Code Camp},
  nobreak=true
  ]
Together with my supervisor and Futhark designer Troels Henriksen,
I was invited for an hour-long presentation and discussion of \futhark{} at the
yearly \textit{SasView Code Camp}, held at the Copenhagen Bio Science Park.\\\\
There was about 20 \sasview{} users and developers attending at the presentation,
which was about the definition and implementation of \futhark{}, along with
a presentation of the result performance boosts for selected models.\\\\
The project was well-received by the audience, and there was an eager discussion
of whether and how it could be possible to really exploit the Futhark
optimizations.\\\\
The primary suggestion was to leverage Futhark to find best-fitting models,
as described in \ref{sec:best-match}, but bundling continuous input from
a sensor to exploit Futhark speed ups at high n was also a feature that many
of the attendants expressed an interest in.
\end{mdframed}

@@

\begin{figure}
  \lstinputlisting[linerange={13-53}]{../sasmodels/sasmodels/models/futharks/header/for_iq.fut}
  \label{fig:kernelfut_loop}
  \caption{The Futhark kernel calculation loop, aping fig \ref{fig:kernelpy_loop}}
\end{figure}
\begin{figure}
  \lstinputlisting[linerange={541-588}, language=Python]{../sasmodels/sasmodels/kernelcl.py}
  \label{fig:kernelcl_call}
  \caption{Calling OpenCL kernel}
\end{figure}

@@

\begin{thebibliography}{1}

\bibitem{futhark}
  T. Henriksen,
  \textit{Futhark performances}
  https://futhark-lang.org/performance.html

\bibitem{pldi17}
  T. Henriksen,
  N. G. W. Serup,
  M. Elsman,
  F. Henglein,
  C. E. Oancea,
  \textit{Futhark: Purely Functional GPU-Programming with Nested Parallelism 
  and In-Place Array Updates}
  Department of Computer Science,
  University of Copenhagen (Denmark),
  2017.

\end{thebibliography}
\end{document}
